---
output:
  github_document:
    html_preview: false
---

# mimar

`mimar` implements a compact chained-imputation workflow in R for missing-data
analysis, artificial amputation, native and learner-backed single and multiple
imputation, diagnostic evaluation, and post-imputation pooling.

The package is built around a complete missing-data workflow: describe the
missingness, create benchmark amputations when needed, impute with native or
learner-backed update rules, inspect diagnostics, evaluate recovered cells when
truth is available, and pool post-fit quantities. The goal is a concise grammar
for the whole workflow, not a replacement for every specialist feature in
larger imputation systems.

The package owns the imputation loop. Every imputer, whether implemented
natively or backed by a learner package, is called the same way:

```r
impute(data, imputer = "pmm", m = 5, maxit = 5, seed = 1)
impute(data, imputer = "rf", m = 5, seed = 1)
impute(data, imputer = "xgboost", m = 5, seed = 1)
```

There is no dependency on `funcml`. Learner-backed imputers call their original
packages directly, and those backend packages are hard dependencies so users can
run any registered imputer without manually resolving learner installations.

## Installation

Install the released version from CRAN:

```r
install.packages("mimar")
```

The CRAN package page is <https://CRAN.R-project.org/package=mimar>.

You can install the development version from GitHub:

```r
install.packages("remotes")
remotes::install_github("ielbadisy/mimar")
```

Then load the package:

```r
library(mimar)
```

## Quick use

For normal use, `impute()` is the only function you need. The input data can
contain `NA`, and the completed outputs returned by `complete()` do not.
Set `verbose = TRUE` when you want a concise progress log for the chained
imputation workflow.

```r
i <- impute(a, imputer = "knn", m = 5, maxit = 5, seed = 1)
complete(i, 1)
complete(i, "all")
```

## Grammar

```r
describe()
ampute()
imputer_registry()
imputer()
impute()
complete()
evaluate()
pool()
plot()
```

## Short example

```{r}

library(mimar)

set.seed(1)
dat <- data.frame(
  age = rnorm(120, 50, 10),
  bmi = rnorm(120, 25, 4),
  sex = factor(sample(c("F", "M"), 120, TRUE)),
  group = factor(sample(c("A", "B", "C"), 120, TRUE)),
  smoker = sample(c(TRUE, FALSE), 120, TRUE)
)

a <- ampute(
  dat,
  prop = 0.25,
  mechanism = "MAR",
  target = c("bmi", "group"),
  by = c("age", "sex"),
  seed = 1
)

i <- impute(a, imputer = "knn", m = 5, maxit = 5, seed = 1, ncore = 2)
complete(i, 1)
summary(i)
evaluate(i)
plot(i, type = "density")
```

## Imputers

Inspect available imputers with:

```{r}
imputer_registry()
describe("imputers")
```

Core native imputers:

- `mean`, `median`, `mode`
- `naive`: median/mode chained baseline
- `norm`: linear normal draw
- `pmm`, `spmm`: predictive mean matching
- `logreg`: binary logistic regression draw
- `polyreg`: one-vs-rest multinomial draw
- `knn`: nearest-neighbor donor imputation
- `hotdeck`: stochastic donor imputation

Learner-backed imputers:

- `rf`: MissForest-style chained random forest imputer through `ranger`
- `ranger`: random forest through `ranger`
- `rpart`: tree imputer through `rpart`
- `nbayes`: naive Bayes through `naivebayes`
- `svm`: support vector machine through `e1071`
- `bart`: Bayesian additive regression trees through `BART`
- `glmnet`: penalized regression through `glmnet`
- `gbm`: gradient boosting through `gbm`
- `xgboost`: gradient boosted trees through `xgboost`
- `famd`: FAMD-assisted donor imputation through `missMDA`
- `superlearner`, `sl`: cross-validated Super Learner-style ensemble imputer

Imputer names are strict: use the names shown by `imputer_registry()`.
Learner-backed imputers are applied as requested to numeric, binary, and
multiclass targets; `mimar` does not silently swap them for another imputer
inside benchmark runs.

## Parallel imputation

The `ncore` argument runs independent completed datasets in parallel. The
parallel boundary is the outer imputation index: each completed dataset gets a
deterministic seed offset, so a fixed `seed`, `m`, `maxit`, and imputer remain
reproducible.

```{r}
i <- impute(a, imputer = "knn", m = 5, maxit = 5, seed = 1, ncore = 2)
```

Use `ncore = 1` for sequential execution, small examples, and the most
conservative behavior in constrained environments.

## Tuning imputers

Learner-backed imputers expose their hyperparameters through `imputer()` or
directly through `...` in `impute()`. Donor-based imputers use the explicit
`donors` argument.

```{r}
rf_spec <- imputer("rf", num.trees = 500)
xgb_spec <- imputer("xgboost", nrounds = 100, max_depth = 3)

i1 <- impute(a, imputer = rf_spec, m = 5, maxit = 5, seed = 1)
i2 <- impute(a, imputer = "xgboost", m = 5, maxit = 5, seed = 1,
             nrounds = 100, max_depth = 3)
i3 <- impute(a, imputer = "knn", m = 5, maxit = 5, seed = 1, donors = 10)
```

The same hyperparameter set is reused across all incomplete variables that a
given imputer supports, which keeps the full chained-imputation pipeline
reproducible and easy to tune.

## Super Learner imputation

`superlearner` combines candidate imputers by cross-validating them on observed
cells, assigning non-negative loss-based weights, and using the weighted
ensemble inside the chained-imputation loop.

```{r}
sl <- imputer(
  "superlearner",
  library = c("pmm", "knn", "rpart"),
  folds = 5,
  metalearner = "inverse_loss"
)

i_sl <- impute(a, imputer = sl, m = 5, maxit = 5, seed = 1)
```

The short alias `imputer = "sl"` is equivalent to `imputer = "superlearner"`.

## Diagnostic Plots

`plot()` methods return `ggplot` objects. For `mimar_imputation` objects, the
main diagnostic types are:

```{r}
plot(i)                                      # imputed cell counts
plot(i, type = "missing")                   # observed/imputed cell map
plot(i, type = "trace", statistic = "mean") # convergence-screening trace
plot(i, type = "density", variable = "bmi") # line-only density overlays
plot(i, type = "boxplot", variable = "bmi") # observed vs imputation 1:m
plot(i, type = "strip", variable = "bmi")   # individual values by imputation
```

Formula diagnostics are available for bivariate and categorical checks:

```{r}
plot(i, type = "xy", formula = bmi ~ age | sex)
plot(i, type = "proportion", formula = group ~ sex)
```

For `type = "xy"`, formulas use `y ~ x` or `y ~ x | group`. For
`type = "proportion"`, formulas use `categorical_variable ~ strata`. Density
diagnostics use line-only overlays so several imputations remain visible rather
than obscuring each other with filled areas.

## Chained Imputation Model

Let `X` be an `n x p` data frame and let `R_ij = 1` when cell `(i, j)` is
missing. For each incomplete variable `X_j`:

- `O_j = {i : R_ij = 0}` are the observed rows
- `M_j = {i : R_ij = 1}` are the missing rows

At each chained update, `mimar` fits an imputer-specific model from the
observed rows and then predicts the missing rows from the current completed
data. In compact form:

```text
fit model on X_-j, O_j -> X_j, O_j
update X_j, M_j using the fitted model
```

Multiple imputation repeats the same chained procedure `m` times with
controlled seeds, bootstrap samples of observed rows, and stochastic prediction
where supported.

Learner-backed imputers are practical stochastic update rules inside this
chained workflow. They can improve predictive recovery, but users should still
inspect trace, distribution, categorical-proportion, and downstream sensitivity
diagnostics rather than assuming every learner automatically supplies proper
multiple-imputation uncertainty for every analysis.

## Algorithm

```text
Input: X, R, h, m, T
Initialize: X~(0) <- init(X)
For k = 1,...,m:
  X~_k(0) <- X~(0)
  For t = 1,...,T:
    For each incomplete variable j:
      B_j <- bootstrap sample of O_j
      fit h on X~_k, B_j, -j and X_Bj,j
      update missing rows M_j using the fitted model
      restore observed rows O_j to their original values
Return: {X~_1(T), ..., X~_m(T)}
```

## Evaluation

When imputation is run on an `ampute()` object, `evaluate()` uses the retained
truth and scores only artificially removed cells. Numeric recovery reports
RMSE, MAE, bias, and correlation. Categorical recovery reports accuracy and
balanced accuracy.

## Pooling

`pool()` combines post-fit quantities estimated separately in each completed
dataset. The statistical target is the quantity itself, not a data frame. A
quantity can be a scalar, coefficient vector, covariance-aware parameter vector,
matrix of survival probabilities, or a scalar metric. Data frames are accepted
only as a tidy adapter for scalar model output.

For survival-probability matrices, `pool_survmat()` applies the complementary
log-log transform internally, pools on that scale, and back-transforms the
result.

For scalar quantities with complete-data variance estimates, `pool()` applies
Rubin-style pooling:

```text
Q_bar = mean(Q_k)
U_bar = mean(U_k)
B     = sample variance of Q_k
T     = U_bar + (1 + 1/m) * B
```

```{r}
results <- data.frame(
  term = rep(c("age", "bmi"), each = 3),
  estimate = c(0.10, 0.11, 0.09, 0.30, 0.32, 0.29),
  std.error = c(0.04, 0.05, 0.04, 0.08, 0.09, 0.08),
  imputation = rep(1:3, times = 2)
)

pool(results)
```

Direct quantity inputs are preferred when available:

```{r}
pool(c(0.10, 0.11, 0.09), std.error = c(0.04, 0.05, 0.04), name = "age")

betas <- list(
  c(age = 0.10, bmi = 0.30),
  c(age = 0.11, bmi = 0.32),
  c(age = 0.09, bmi = 0.29)
)
covariances <- list(
  diag(c(0.04, 0.08)^2),
  diag(c(0.05, 0.09)^2),
  diag(c(0.04, 0.08)^2)
)
pool(betas, covariance = covariances)
```

When no reliable complete-data variance is supplied, as is common for some
performance metrics, `pool()` reports robust summaries by default: median,
interquartile range, and range across imputations.

## Installation notes

Learner backends are hard dependencies. Installing `mimar` installs the
packages needed by the registered learner-backed imputers, including `ranger`,
`rpart`, `naivebayes`, `e1071`, `BART`, `glmnet`, `gbm`, `xgboost`, and
`missMDA`.
