---
title: "Compact Multiple Imputation Workflows with mimar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compact Multiple Imputation Workflows with mimar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

`mimar` exposes a compact grammar for missing-data description, artificial missingness, imputation, evaluation, pooling, and plotting:

```r
describe()
ampute()
impute()
evaluate()
pool()
plot()
```

The package exposes missing-data verbs plus a small imputer learner contract. Every imputer has the same standard lifecycle: construct it with `imputer()`, fit it with `fit()`, and generate imputations with `predict()`. The chained-equation imputation engine is selected with one public argument:

```r
impute(data, m = 5, imputer = "mi")
```

Supported imputer names include classical methods such as `"pmm"`, `"norm"`, `"logreg"`, and `"polyreg"`, plus ML learners such as `"randomForest"`, `"knn"`, `"xgboost"`, `"svm"`, `"bart"`, `"naive_bayes"`, `"rpart"`, and `"glmnet"`.

Let \(X\) be an \(n \times p\) data matrix with entries \(X_{ij}\), where \(i = 1,\ldots,n\) indexes rows and \(j = 1,\ldots,p\) indexes variables. Define the missingness indicator

\[
R_{ij} =
\begin{cases}
1, & X_{ij}\ \text{is missing},\\
0, & X_{ij}\ \text{is observed}.
\end{cases}
\]

`mimar` uses three masks internally:

* `mask_original`: cells already missing before artificial amputation;
* `mask_added`: cells newly made missing by `ampute()`;
* `mask_total`: all missing cells after amputation.

When known truth is available, recovery is evaluated only where `mask_added == TRUE`. Naturally missing cells are preserved and not scored against truth.

```{r}
library(mimar)
set.seed(1)
dat <- data.frame(
  age = rnorm(120, 50, 10),
  bmi = rnorm(120, 25, 4),
  sex = factor(sample(c("F", "M"), 120, TRUE)),
  group = factor(sample(c("A", "B", "C"), 120, TRUE)),
  smoker = sample(c(TRUE, FALSE), 120, TRUE)
)

head(dat)
```

## Missingness Description

`describe()` computes variable summaries, row summaries, missingness patterns, and complete-case counts. For variable \(j\), the missing proportion is

\[
\hat{\pi}_j = \frac{1}{n}\sum_{i=1}^{n} R_{ij}.
\]

The complete-case count is

\[
n_{\mathrm{cc}} = \sum_{i=1}^{n} I(R_{i1}=0,\ldots,R_{ip}=0).
\]

```{r}
d <- describe(dat)
d
summary(d)
head(d$row_summary)
```

```{r, fig.width=7, fig.height=4.5}
plot(d)
```

`describe("imputers")` lists the compact imputer names that can be passed to `impute()`.

```{r}
available_imputers <- describe("imputers")
available_imputers
```

```{r, fig.width=7, fig.height=4.5}
plot(available_imputers)
```

## Missingness Mechanisms and Amputation

`ampute()` introduces artificial missingness for simulation, benchmarking, teaching, and sensitivity analysis. It does not prove or infer the real-world mechanism for an observed dataset.

### MCAR

Missing completely at random is defined as

\[
P(R_j = 1 \mid X_{\mathrm{obs}}, X_{\mathrm{mis}}) = P(R_j = 1).
\]

In `mimar`, MCAR missingness is generated independently of observed and unobserved values. Existing missing cells are preserved, and new missingness is added only among observed cells.

### MAR

Missing at random is defined as

\[
P(R_j = 1 \mid X_{\mathrm{obs}}, X_{\mathrm{mis}})
= P(R_j = 1 \mid X_{\mathrm{obs}}).
\]

For MAR amputation, missingness in `target` depends on observed variables in `by`. The target variable itself must not be included in `by`. `mimar` constructs a standardized observed predictor score \(s_i\) and uses

\[
\eta_i = \alpha + \beta s_i,
\qquad
p_i = \operatorname{logit}^{-1}(\eta_i)
= \frac{1}{1 + \exp(-\eta_i)}.
\]

The intercept \(\alpha\) is calibrated so that

\[
\frac{1}{n}\sum_{i=1}^{n} p_i \approx \texttt{prop}.
\]

### MNAR

Missing not at random is defined as

\[
P(R_j = 1 \mid X_{\mathrm{obs}}, X_{\mathrm{mis}})
= P(R_j = 1 \mid X_j).
\]

For numeric variables, `direction = "right"` makes larger values more likely missing, `direction = "left"` makes smaller values more likely missing, and `direction = "both"` makes extreme values more likely missing. For categorical variables, missingness may depend on category patterns or level frequencies.

```{r}
a_mcar <- ampute(dat, prop = 0.1, mechanism = "MCAR", seed = 1)
a_mar <- ampute(dat, prop = 0.25, mechanism = "MAR",
                target = c("bmi", "group"), by = c("age", "sex"), seed = 1)
a_mnar <- ampute(dat, prop = 0.2, mechanism = "MNAR",
                 target = "bmi", direction = "right", seed = 1)

a_mar
summary(a_mar)
head(a_mar$mask_added)
```

```{r, fig.width=7, fig.height=4.5}
plot(a_mar)
```

## Single and Multiple Imputation

`impute()` always returns a `mimar_imputation` object. The completed datasets are the primary payload: use `complete()` to extract one completed dataset, all completed datasets, or a stacked long data frame.

For `"meanmode"`, numeric and integer variables are imputed with the median:

\[
\tilde{x}_j = \operatorname{median}\{x_{ij}: R_{ij}=0\}.
\]

Logical, factor, and character variables are imputed with the mode:

\[
\tilde{x}_j = \arg\max_c \sum_{i=1}^{n} I(x_{ij}=c,\ R_{ij}=0).
\]

This imputer is deterministic, so multiple completed datasets are identical and `stochastic = FALSE`.

```{r}
i1 <- impute(a_mar, m = 1, imputer = "meanmode", seed = 1)
i5 <- impute(a_mar, m = 5, imputer = "meanmode", seed = 1)

i1
i5
complete(i1)
complete(i5, "long")
summary(i5)
length(i5$imputations)
anyNA(i5$imputations[[1]])
```

```{r, fig.width=7, fig.height=4.5}
plot(i5)
plot(i5, type = "missing")
plot(i5, type = "density")
plot(i5, type = "methods")
```

## Imputer Learners

`describe("imputers")` lists the standalone imputer learners. Classical learners are implemented in `mimar`; ML learners use the required `funcml` fit/predict API behind the same `mimar` imputer contract.

Each imputer declares whether it supports numeric, binary, and multiclass targets. If an incompatible imputer is requested for a variable, `mimar` errors instead of falling back silently.

The chained ML imputation algorithm is:

1. initialize missing predictors with the mean/mode imputer;
2. for each incomplete variable \(X_j\), split rows into observed and missing sets;
3. fit the selected learner on \(X_j\) using all other variables \(X_{-j}\) as predictors;
4. predict missing values in \(X_j\);
5. repeat for `maxit` iterations.

Conceptually, for each incomplete variable \(j\), the internal model is

\[
\hat{f}_j: X_{-j} \rightarrow X_j.
\]

The public API still uses only `imputer`:

```{r}
describe("imputers")
```

Use any listed learner through the same compact API:

```r
impute(a_mar, m = 5, imputer = "randomForest", seed = 1)
```

The following chunk runs a lightweight classical chained-equation imputation.

```{r}
i_pmm <- impute(a_mar, m = 1, imputer = "mi", maxit = 1, seed = 1)
i_pmm
evaluate(i_pmm)$summary
```

`mice` is not wrapped. It remains a methodological reference for chained equations, while `mimar` owns the imputation loop and uses standalone imputer learners.

## Evaluation Using Known Truth

`evaluate()` summarizes imputation quality. Without truth, it reports distribution and variability diagnostics. With amputation truth, it also computes recovery metrics on artificially amputed cells.

For numeric variables, with truth \(y_i\) and imputed values \(\hat{y}_i\), the metrics are:

\[
\mathrm{RMSE} =
\sqrt{\frac{1}{k}\sum_{i=1}^{k}(\hat{y}_i-y_i)^2},
\qquad
\mathrm{MAE} =
\frac{1}{k}\sum_{i=1}^{k}|\hat{y}_i-y_i|,
\]

\[
\mathrm{Bias} =
\frac{1}{k}\sum_{i=1}^{k}(\hat{y}_i-y_i),
\qquad
\mathrm{Cor} = \operatorname{cor}(y,\hat{y}).
\]

For categorical variables, accuracy is

\[
\mathrm{Accuracy} =
\frac{1}{k}\sum_{i=1}^{k} I(\hat{y}_i = y_i).
\]

Balanced accuracy is the mean class-specific recall:

\[
\mathrm{BalancedAccuracy}
= \frac{1}{C}\sum_{c=1}^{C}
\frac{\sum_i I(y_i=c,\ \hat{y}_i=c)}{\sum_i I(y_i=c)}.
\]

```{r}
e <- evaluate(i5)
e
e$recovery
e$variability
```

```{r, fig.width=7, fig.height=4.5}
plot(e)
```

## Pooling External Results

`mimar` does not fit user models. Supply external results and pool them.

For scalar estimates \(q_1,\ldots,q_m\) with variances \(u_i = \mathrm{SE}(q_i)^2\), Rubin's rules are:

\[
\bar{q} = \frac{1}{m}\sum_{i=1}^{m} q_i,
\qquad
\bar{u} = \frac{1}{m}\sum_{i=1}^{m} u_i,
\]

\[
b = \frac{1}{m-1}\sum_{i=1}^{m}(q_i-\bar{q})^2,
\qquad
t = \bar{u} + \left(1+\frac{1}{m}\right)b,
\]

\[
\mathrm{SE}_{\mathrm{pooled}} = \sqrt{t}.
\]

```{r}
external_results <- data.frame(
  term = rep(c("age", "bmi"), each = 5),
  estimate = c(0.10, 0.11, 0.09, 0.12, 0.10, 0.30, 0.32, 0.29, 0.31, 0.33),
  std.error = c(0.04, 0.05, 0.04, 0.05, 0.04, 0.08, 0.09, 0.08, 0.09, 0.08),
  imputation = rep(1:5, times = 2)
)
p <- pool(external_results)
p
```

```{r, fig.width=6, fig.height=4}
plot(p)
```

Metrics without standard errors are pooled by their mean and between-imputation variability:

\[
\bar{m} = \frac{1}{M}\sum_{r=1}^{M} m_r,
\qquad
s_m^2 = \frac{1}{M-1}\sum_{r=1}^{M}(m_r-\bar{m})^2.
\]

```{r}
metrics <- data.frame(
  metric = rep(c("auc", "rmse"), each = 5),
  value = c(0.81, 0.83, 0.80, 0.82, 0.84, 1.2, 1.1, 1.3, 1.2, 1.0),
  imputation = rep(1:5, times = 2)
)

pool(metrics)
```

## Plotting

Each core object has a plot method:

```{r, fig.width=7, fig.height=4.5}
plot(describe(dat))
plot(a_mar)
plot(i5)
plot(e)
plot(p)
plot(describe("imputers"))
```

Imputation objects expose diagnostic plot types for missingness maps, observed-vs-imputed distributions, imputation methods, and between-imputation variability.

```{r, eval=requireNamespace("ggplot2", quietly = TRUE), fig.width=7, fig.height=4.5}
plot(describe(a_mar$data), engine = "ggplot2")
plot(i5, type = "strip")
plot(i5, type = "variability")
```

## Object Classes

The returned objects are stable S3 classes:

```{r}
sapply(
  list(
    missing = d,
    amputation = a_mar,
    imputation = i5,
    evaluation = e,
    pool = p,
    imputers = available_imputers
  ),
  function(x) class(x)[1]
)
```

These objects are ordinary lists or data frames with stable fields, so they can be inspected directly.

## Limitations

The baseline `"meanmode"` imputer supports numeric, integer, logical, factor, character, and Date variables. Unsupported variable types are left unchanged with a warning.

ML learners are provided through `funcml`, which is a package dependency. The user-facing object remains a `mimar` imputer with the same `fit()` and `predict()` behavior as classical learners.

Missingness mechanisms are data-generating assumptions. Observed data alone cannot prove that real-world missingness is MCAR, MAR, or MNAR.
