---
title: "Chained Multiple Imputation Workflows with mimar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chained Multiple Imputation Workflows with mimar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Purpose

`mimar` provides one grammar for missing-data workflows:

```r
describe()
ampute()
imputer_registry()
impute()
complete()
evaluate()
pool()
plot()
```

The imputation principle is simple: `mimar` owns the chained imputation
procedure, and every imputer is a variable-update rule inside that procedure.
Native imputers and optional learner-backed imputers are called the same way:

```r
impute(data, imputer = "pmm", m = 5, maxit = 5)
impute(data, imputer = "rf", m = 5, maxit = 5)
impute(data, imputer = "xgboost", m = 5, maxit = 5)
```

There is no dependency on `funcml`; wrapped learners call their original
packages directly when requested.

```{r}
library(mimar)

set.seed(1)
dat <- data.frame(
  age = rnorm(120, 50, 10),
  bmi = rnorm(120, 25, 4),
  sex = factor(sample(c("F", "M"), 120, TRUE)),
  group = factor(sample(c("A", "B", "C"), 120, TRUE)),
  smoker = sample(c(TRUE, FALSE), 120, TRUE)
)

head(dat)
```

## Missingness Notation

Let \(X\) be an \(n \times p\) data matrix with entries \(X_{ij}\). Define the
missingness indicator

\[
R_{ij} =
\begin{cases}
1, & X_{ij}\ \text{is missing},\\
0, & X_{ij}\ \text{is observed}.
\end{cases}
\]

For variable \(j\), the missingness proportion is

\[
\hat{\pi}_j = \frac{1}{n}\sum_{i=1}^{n} R_{ij}.
\]

The complete-case count is

\[
n_{\mathrm{cc}} = \sum_{i=1}^{n} I(R_{i1}=0,\ldots,R_{ip}=0).
\]

```{r}
d <- describe(dat)
d
summary(d)
```

```{r, fig.width=7, fig.height=4}
plot(d)
```

## Imputer Registry

`imputer_registry()` is the exposed registry. It reports each imputer, whether
it is implemented natively or through an optional package, the package required
when relevant, and target-type support.

```{r}
imputer_registry()
```

`describe("imputers")` returns the same registry with the `mimar_imputers`
print and plot methods.

```{r}
describe("imputers")
```

## Artificial Missingness

`ampute()` creates benchmark datasets and keeps three masks:

* `mask_original`: cells missing before amputation;
* `mask_added`: cells made missing by `ampute()`;
* `mask_total`: all cells missing after amputation.

For a target variable \(X_j\), new missingness is generated by

\[
A_{ij} \sim \operatorname{Bernoulli}(p_i),
\qquad
X_{ij} \leftarrow \mathrm{NA}\ \text{if}\ A_{ij}=1.
\]

### MCAR

For missing completely at random,

\[
p_i = \pi,
\]

where \(\pi\) is `prop`.

### MAR

For missing at random, `mimar` builds a standardized score \(s_i\) from the
observed `by` variables and uses

\[
p_i = \operatorname{logit}^{-1}(\alpha + s_i)
= \frac{1}{1 + \exp[-(\alpha+s_i)]}.
\]

The intercept \(\alpha\) is calibrated so that

\[
\frac{1}{n}\sum_{i=1}^{n} p_i \approx \texttt{prop}.
\]

### MNAR

For missing not at random, \(s_i\) is derived from the target itself. Numeric
targets can emphasize the right tail, left tail, or both tails.

```{r}
a <- ampute(
  dat,
  prop = 0.25,
  mechanism = "MAR",
  target = c("bmi", "group"),
  by = c("age", "sex"),
  seed = 1
)

a
summary(a)
```

## Chained Imputation Algorithm

For incomplete variable \(X_j\), define

\[
O_j = \{i: R_{ij}=0\},
\qquad
M_j = \{i: R_{ij}=1\}.
\]

At iteration \(t\), with the current completed data \(\tilde X^{(t)}\), `mimar`
fits

\[
\hat f_j^{(t)}:
\tilde X_{-j,O_j}^{(t)} \rightarrow X_{j,O_j}.
\]

Then it updates the missing cells:

\[
\tilde X_{j,M_j}^{(t+1)}
= g_j\left(\hat f_j^{(t)}, \tilde X_{-j,M_j}^{(t)}\right),
\]

where \(g_j\) is the prediction or donor-sampling rule for the selected imputer.

### Pseudo Algorithm

```text
Input:
  X: data frame with missing values
  h: imputer name or named imputer vector
  m: number of completed datasets
  T: number of chained iterations

Initialize:
  Fill missing numeric values with medians.
  Fill missing categorical values with modes.

For k = 1,...,m:
  Set seed for imputation k.
  Start from initialized data X_tilde.

  For t = 1,...,T:
    For each variable j with missing cells:
      O_j <- observed rows for X_j
      M_j <- missing rows for X_j
      h_j <- selected imputer for X_j
      B_j <- bootstrap sample from O_j

      Fit h_j on:
        predictors X_tilde[B_j, -j]
        target     X[B_j, j]

      Predict or sample:
        X_tilde[M_j, j] <- h_j(X_tilde[M_j, -j])

      Restore storage type:
        numeric, integer, Date, logical, factor, or character

  Store X_tilde as completed dataset k.

Return:
  mimar_imputation object with m completed datasets and diagnostics.
```

## Implemented Update Equations

### Mean, Median, Mode, and Naive

For numeric targets, `mean` and `median` use

\[
\tilde x_{ij} = \bar x_j
\quad\text{or}\quad
\tilde x_{ij} = \operatorname{median}(x_{O_j,j}).
\]

For categorical targets, `mode` uses

\[
\tilde x_{ij}
= \arg\max_c \sum_{i \in O_j} I(x_{ij}=c).
\]

`naive` uses median for numeric-like variables and mode for categorical-like
variables.

### Normal Linear Draw

`norm` fits a linear model

\[
X_j = X_{-j}\beta_j + \epsilon_j,
\qquad
\epsilon_j \sim N(0,\sigma_j^2).
\]

For missing rows,

\[
\tilde X_{j,M_j} = \hat X_{-j,M_j}\hat\beta_j + e,
\qquad
e \sim N(0,\hat\sigma_j^2).
\]

### Predictive Mean Matching

`pmm` fits the same linear model and computes predicted means
\(\hat\mu_i\). For each missing row \(i\), donor candidates are the observed
rows with smallest distance

\[
d(i,\ell) = |\hat\mu_i - \hat\mu_\ell|.
\]

The imputed value is sampled from the nearest donor set:

\[
\tilde x_{ij} = x_{\ell j}, \qquad \ell \sim \mathcal{D}_i.
\]

`spmm` uses the same update rule but is treated as a single-step PMM method.

### Logistic Regression

For binary targets, `logreg` fits

\[
\Pr(X_j = 1 \mid X_{-j}) =
\operatorname{logit}^{-1}(X_{-j}\beta_j).
\]

Missing classes are drawn from

\[
\tilde X_{ij} \sim \operatorname{Bernoulli}(\hat p_i).
\]

### One-vs-Rest Polynomial Regression

For multiclass targets \(c=1,\ldots,C\), `polyreg` fits one-vs-rest logistic
models

\[
\Pr(X_j=c \mid X_{-j}) =
\operatorname{logit}^{-1}(X_{-j}\beta_{jc}).
\]

Probabilities are normalized across classes:

\[
\tilde p_{ic}
= \frac{\hat p_{ic}}{\sum_{r=1}^{C}\hat p_{ir}},
\]

and missing classes are sampled from \(\operatorname{Categorical}(\tilde p_i)\).

### Donor Methods: knn, hotdeck, famd

`knn` and `hotdeck` are donor-update rules. Predictors are encoded into a
numeric design matrix \(Z\). For a missing row \(i\), distance to observed row
\(\ell\) is

\[
d(i,\ell) = \left\|Z_i - Z_\ell\right\|_2.
\]

`knn` samples from the nearest donor set. `hotdeck` uses the same standardized
donor mechanism as a stochastic hot-deck update inside the chain. `famd` first
builds FAMD-assisted completed predictor coordinates when `missMDA` is
available, then uses donor matching in that coordinate space.

### Learner-Backed Imputers

`rf`, `ranger`, `rpart`, `naive_bayes`, `svm`, `bart`, `glmnet`, `gbm`, and
`xgboost` are fitted directly as variable-level supervised learners:

\[
\hat f_j = \mathcal{A}_h(X_{-j,O_j}, X_{j,O_j}),
\]

where \(\mathcal{A}_h\) is the selected learner. Numeric targets use regression
predictions. Binary and multiclass targets use class predictions or sampled
class probabilities when the backend exposes probabilities.

`rf` is a random forest imputer inside `mimar`'s chained update loop. It is a
variable-level supervised update, not a whole-dataframe imputation engine.

## Single and Multiple Imputation

```{r}
i_naive <- impute(a, imputer = "naive", m = 3, maxit = 3, seed = 1)
i_naive
summary(i_naive)
complete(i_naive, 1)
```

All imputers use the same public call:

```{r}
i_knn <- impute(a, imputer = "knn", m = 1, maxit = 2, seed = 1)
i_hotdeck <- impute(a, imputer = "hotdeck", m = 1, maxit = 2, seed = 1)

summary(i_knn)
summary(i_hotdeck)
```

Optional learner-backed imputers can be used when the backend package is
installed:

```r
impute(a, imputer = "rf", m = 5, maxit = 5, seed = 1)
impute(a, imputer = "glmnet", m = 5, maxit = 5, seed = 1)
impute(a, imputer = "xgboost", m = 5, maxit = 5, seed = 1)
```

## Evaluation

When an imputation object contains amputation truth, `evaluate()` scores only
cells where `mask_added == TRUE`.

For numeric truth \(y_i\) and imputed values \(\hat y_i\),

\[
\operatorname{RMSE} =
\sqrt{\frac{1}{q}\sum_{i=1}^{q}(\hat y_i-y_i)^2},
\]

\[
\operatorname{MAE} =
\frac{1}{q}\sum_{i=1}^{q}|\hat y_i-y_i|,
\qquad
\operatorname{Bias} =
\frac{1}{q}\sum_{i=1}^{q}(\hat y_i-y_i).
\]

For categorical targets,

\[
\operatorname{Accuracy}
= \frac{1}{q}\sum_{i=1}^{q} I(\hat y_i=y_i).
\]

Balanced accuracy averages per-class recall.

```{r}
e <- evaluate(i_naive)
e
describe(e)
head(e$recovery_by_imputation)
```

```{r, fig.width=7, fig.height=4}
plot(i_naive)
plot(i_naive, type = "missing")
plot(i_naive, type = "density")
plot(e)
```

## Pooling

For estimates \(Q_k\) and complete-data variances \(U_k\) from completed
dataset \(k\),

\[
\bar Q = \frac{1}{m}\sum_{k=1}^{m}Q_k,
\qquad
\bar U = \frac{1}{m}\sum_{k=1}^{m}U_k,
\]

\[
B = \frac{1}{m-1}\sum_{k=1}^{m}(Q_k-\bar Q)^2,
\qquad
T = \bar U + \left(1+\frac{1}{m}\right)B.
\]

`pool()` returns pooled estimates, total standard errors, confidence intervals,
and missing-information diagnostics.

```{r}
external_results <- data.frame(
  term = rep(c("age", "bmi"), each = 3),
  estimate = c(0.10, 0.11, 0.09, 0.30, 0.32, 0.29),
  std.error = c(0.04, 0.05, 0.04, 0.08, 0.09, 0.08),
  imputation = rep(1:3, times = 2)
)

p <- pool(external_results)
p
```
