---
title: "mimar simulation report"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggplot2)
library(tibble)
library(knitr)
```

```{r load}
res <- readRDS("analysis/simulation/results/simulation_results.rds")
results <- as.data.frame(res$results)
summary_stats <- as.data.frame(res$summary)
ranking <- as.data.frame(res$ranking)
failure_rate <- as.data.frame(res$failure_rate)
params <- res$params
```

## Design

This simulation varies MNAR missingness from 0.1 to 0.7, uses `m = 5` completed
datasets and `maxit = 5` chained iterations, and repeats each condition across
`r params$n_rep` Monte Carlo replicates. The data-generating processes are:

1. linear Gaussian dependence
2. nonlinear interaction structure
3. latent factor structure

## Summary

```{r}
kable(head(ranking, 12), digits = 3)
```

## Main Plot

```{r fig.width=14, fig.height=10}
plot_data <- subset(summary_stats, status == "ok")
plot_data$method <- factor(plot_data$method, levels = unique(ranking$method))

ggplot(plot_data, aes(x = prop, y = nrmse, color = dgp, group = dgp)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  facet_wrap(~ method, ncol = 4) +
  scale_x_continuous(breaks = seq(0.1, 0.7, by = 0.1)) +
  labs(
    x = "MNAR proportion",
    y = "Mean NRMSE",
    color = "DGP",
    title = "MNAR recovery curves by imputer"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

## Failure Rate

```{r fig.width=14, fig.height=10}
fail_plot <- failure_rate
fail_plot$method <- factor(fail_plot$method, levels = unique(ranking$method))
ggplot(fail_plot, aes(x = prop, y = failure_rate, color = dgp, group = dgp)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  facet_wrap(~ method, ncol = 4) +
  scale_x_continuous(breaks = seq(0.1, 0.7, by = 0.1)) +
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", 100 * x)) +
  labs(
    x = "MNAR proportion",
    y = "Failure rate",
    color = "DGP",
    title = "Imputation failures by imputer"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

## Notes

The ranking table orders methods by mean NRMSE and then by runtime. In the full
paper draft, these outputs can be used to build a main comparison table and a
supplementary figure on runtime and failure behavior.
